# Reproducible Research: Peer Assessment 1



## Loading and preprocessing the data

```{r}
library(knitr)
data <- read.csv(unz("activity.zip", "activity.csv"), header=T, quote="\"",
                 sep=",")
data$date = as.Date(data$date, format = "%Y-%m-%d")

# Convert the interval to time
paddedInterval <- paste0("0000", data$interval)
mins <- substr(paddedInterval, nchar(paddedInterval) - 1, 
               nchar(paddedInterval))
hrs <- substr(paddedInterval, nchar(paddedInterval) - 3, 
              nchar(paddedInterval) - 2)
data$time <- paste0(hrs, ":", mins)

# Store if the date is weekend/weekday for later use
data$dayType <- factor(ifelse(weekdays(data$date) %in% c("Saturday",
                                                               "Sunday"),
                                 "Weekend", "Weekday"))

completeData <- data[complete.cases(data),]
```



## What is mean total number of steps taken per day?

First, we take the steps and calculate their sums by date.

```{r}
library(ggplot2)

totalStepsPerDay <- tapply(completeData$steps, completeData$date, 
                           sum, na.rm=TRUE)
```

Now, we calculate the mean and median of those totals.

```{r}
meanStepsPerDay <- mean(totalStepsPerDay)
medianStepsPerDay <- median(totalStepsPerDay)

```

First, we'll plot the total number of steps taken, by day, as a simple histogram.

```{r}
ggplot(completeData, aes(x=date, y=steps)) + 
    geom_histogram(stat="identity", fill = "blue") +
    labs(title="Histogram of Total Number of Steps by Day") +
    labs(x = "Day", y = "Steps")


```

Next, let's look at a histogram of the total number of steps per day. In magenta,
we have the mean steps taken per day. In blue, the median steps taken per day.

```{r}

hist(totalStepsPerDay, col="green", breaks=10, 
     main = "Histogram of Total Number of Steps Per Day",
     xlab = "Total Steps Per Day")
abline(v = meanStepsPerDay, col ="magenta", lwd = 4)
abline(v = medianStepsPerDay, col = "blue", lwd=2)
```

The mean steps per day is `r prettyNum(meanStepsPerDay, ",")`. The median steps per day is
`r prettyNum(medianStepsPerDay, ",")`.

## What is the average daily activity pattern?

Calculate the mean steps taken for each interval (averaged across all days).
Then, plot the average.

```{r}
meanStepsByInterval <- tapply(completeData$steps, completeData$time, 
                              mean, na.rm = TRUE)
plot(meanStepsByInterval, axes = FALSE, t = "l", xlab="Interval", 
     ylab = "Average Steps Taken Across All Days")
axis(2)
axis(1, at=seq(1,length(meanStepsByInterval),12), 
     labels=unlist(dimnames(meanStepsByInterval))[seq(1,length(meanStepsByInterval),12)] )
```

```{r}
intervalWithMaxSteps <- names(which(meanStepsByInterval == 
                                        max(meanStepsByInterval)))
```

The interval with the maximum average across all days is `r intervalWithMaxSteps`.

## Imputing missing values

How many missing values are there?

```{r}
totalMissingRows <- sum(!complete.cases(data))
```

There are `r totalMissingRows` missing values in the original data set. We will
replace these missing values with the mean (over all days) for that interval.

```{r}
newData <- data # new data frame to hold imputed data
# Get missing indices for steps
indexForNA <- which(is.na(newData$steps))
for (j in seq_along(indexForNA)) {
    newData[indexForNA[j], "steps"] <- as.integer(meanStepsByInterval[
        as.character(newData[indexForNA[j], "time"])])
}

newTotalMissingRows <- sum(!complete.cases(newData))
```

There are now `r newTotalMissingRows` missing values in the imputed data set.
Now let's take a look at the total steps per day as a histogram, and the new
mean (magenta) and median (blue).

```{r}

newTotalStepsPerDay <- tapply(newData$steps, newData$date, 
                           sum)
newMeanStepsPerDay <- mean(newTotalStepsPerDay)
newMedianStepsPerDay <- median(newTotalStepsPerDay)

hist(newTotalStepsPerDay, col="green", breaks=10, 
     main = "Histogram of Total Number of Steps Per Day",
     xlab = "Total Steps Per Day")
abline(v = newMeanStepsPerDay, col ="magenta", lwd = 4)
abline(v = newMedianStepsPerDay, col = "blue", lwd=2)
```

The mean steps per day is `r prettyNum(newMeanStepsPerDay, ",")`. The median
steps per day is `r prettyNum(newMedianStepsPerDay, ",")`. They are close to the
original, but slightly lower. Overally, the imputing of the data kept the
mean/median close to the originals.

## Are there differences in activity patterns between weekdays and weekends?

```{r}
# Calculate mean steps by dayType and interval
meanStepsByIntervalDayType <- with(newData, tapply(steps, list(dayType, time),
                                                   mean))

par(mfrow = c(2,1), mar = c(2,4,1,0), oma = c(4, 2, 0, 4))
plot(meanStepsByIntervalDayType[1, ], axes = FALSE,t = "l", main = "weekday", ylab = "", xlab = "")
axis(2)
axis(1, at=seq(1,length(meanStepsByIntervalDayType[1, ]),12), 
     labels=unlist(names(meanStepsByIntervalDayType[1, ]))[seq(1,length(meanStepsByIntervalDayType[1, ]),12)] )
plot(meanStepsByIntervalDayType[2, ], axes = FALSE, t = "l", main = "weekend", ylab = "", xlab = "")
axis(2)
axis(1, at=seq(1,length(meanStepsByIntervalDayType[2, ]),12), 
     labels=unlist(names(meanStepsByIntervalDayType[2, ]))[seq(1,length(meanStepsByIntervalDayType[2, ]),12)] )
mtext("Number of Steps", side = 2, outer = TRUE)
mtext("Interval", side=1, outer=TRUE, padj=1)
```

There are a lot more steps taken in the early morning on weekdays than on weekends.
People walk later in the evenings on weekends. More steps are taken on weekends.